BOLD Parameter Estimation Using Particle Filters

### 2010-08-16 doc/thesis/conclusion.tex
spelling fixes etc.

The primary reason this method has not been used before is the high 
dimensionality of the system. In Murray, 2008 the idea of learning
the parameters is floated as the correct method, yet 
learning 7 parameters with a Monte Carlo method was deemed intractable
\cite{Murray2008}. The 
concern was that so many parameters creates a prohibitively large 
search space, which requires too many particle to learn properly. To overcome
this difficulty, instead of starting the algorithm with 1000 particles,
the initial number of particles was set to 16,000. This meant that
when the search space was the largest, the number of particles was
sufficiently dense to represent the prior distribution. The result is
a particle filter algorithm that spends computing resources only where
it is really needed.

The particle filter is a true Bayesian non-parametric algorithm that, given
enough measurements will approach the true probability distribution
of the parameters. The conclusion that the parameters are not
uniquely identifiable is a disappointing one in light of the plethora
of research papers that have attempted to learn them. However, the fact
that the parameters are ill-posed further demonstrates
the need to estimate probability distribution rather than a single 
parameter estimate. At the same time, the output of the particle filter
is still able to give good estimates of the BOLD output, and is not
dependent on heavy Gaussian smoothing. Therefore, all the current 
experimental paradigms to determine regional activity are still viable.
To borrow a computer term, the particle filter algorithm is backward 
compatible with the current methods. Indeed, the particle filter excelled
at determining activation in very low SNR simulations 
(\autoref{sec:Multi-voxel Simulation}). Not only that, but the particle
filter algorithm identified areas of activation that were completely
missed by SPM (\autoref{fig:comp6}).

@@

Concluding, the particle filter provides comparable performance with
conventional tests, but also provides a platform for a wide range of
future uses beyond what is possible with the methods applied in the
past.

>>>>

The primary reason this method has not been used before is the high
dimensionality of the system. In Murray, 2008 the idea of learning
the parameters is floated as the best method, yet
learning 7 parameters with a Monte Carlo method was deemed intractable
\cite{Murray2008}. The
concern was that so many parameters creates a prohibitively large
search space, which requires too many particle to learn properly. To overcome
this difficulty, instead of starting the algorithm with 1000 particles,
the initial number of particles was set to 16,000. This meant that
when the search space was the largest, the number of particles was
sufficiently dense to represent the prior distribution. The result is
a particle filter algorithm that spends computing resources only where
it is really needed.

The particle filter is a Bayesian non-parametric algorithm that, given
enough measurements will approach the true probability distribution
of the parameters. The conclusion that the parameters are not
uniquely identifiable is disappointing in light of the plethora
of research papers that have attempted to learn them. However, the fact
that the parameters are ill-posed further demonstrates
the need to estimate a probability distribution rather than a single
parameter estimate. At the same time, the output of the particle filter
is still able to give good estimates of the BOLD output, and is not
dependent on heavy Gaussian smoothing. Therefore, all the current
experimental paradigms to determine regional activity are still viable.
To borrow a computer term, the particle filter algorithm is backward
compatible with the current methods. Indeed, the particle filter excelled
at determining activation in very low SNR simulations
(\autoref{sec:Multi-voxel Simulation}). Not only that, but the particle
filter algorithm identified areas of activation that were completely
missed by SPM (\autoref{fig:comp6}).

@@

Concluding, the particle filter provides comparable performance with
conventional tests, but also provides a platform for a wide range of
future uses beyond what was possible with the methods that have been
used before.

### 2010-08-16 doc/thesis/discussion.tex
spelling fixes etc.

\chapter{Discussion}
\label{sec:Discussion}
\section{Review of Results}
% Overview of results
%% Parameters under-constrained
%% Output estimates good
There are two possible causes for the large variance in
parameter estimates. First, it could be that the particle filter is
incapable of learning the model. However, given the high quality
of BOLD estimates, this is almost certainly not the case. The other 
possible explanation is that the given stimulus is not sufficiently
varied to bring out all the properties of the system. While it is 
possible that this is the case, it is unlikely that any
stimulus sequence could differentiate all the parameters. 
Its also extremely intesting that the histograms of parameters' mean
approach the width of the prior, even when the particle filter converges
proprerly to far thinner distributions. Its possible the prior 
distribution is far wider than 
it needs to be. Finally, the fact that \autoref{sec:VeryLongSim}, which 
used randomly spaced impulses,
still had a high correlation further indicates that the parameters
are ill-defined.

Thus there is little doubt that the parameters of the BOLD model are under
constrained. While unsurprising given the sensitivity analysis by Deneux et al.,
it is a notable conclusion \cite{Deneux2006}. Other methods
that depend on a point estimate of the parameters, such as least squares 
or Kalman filters (which uses the first two moments) are limited in their
capacity to estimate such parameters. While estimates of
the  BOLD signal may still be correct, the 
underlying parameters and state variables cannot be described without using
a joint probability distribution function. In this sense, particle 
filters represent an important step forward in BOLD parameter 
estimation. Representing uncertainty with a mean
and variance is insufficient; so using a particle filter
or Bayesian estimate of the posterior is not simply an enhancement, 
but a necessary precaution.

@@

As the results in \autoref{sec:RealData} show, the heatmaps, especially
those of mutual information, closely resembled the results of SPM. While the activation
tests were more sensitive, there were some additional false positives, 
though the problem is difficult to quantify. Regions with M.I. above 
$0.15$ consistently fit the FMRI data well, and simulations showed that
the particle filter performed extremely well in the face of significant
noise. In sum, the estimated BOLD output remained consistent despite
large swings in parameters.

@@

A more practical benefit with the particle filter is that it is mathematically
simple. An understanding of Bayesian statistics is all that is necessary to understand
how the particle filter works.  Additionally very few assumptions
are needed for the particle filter. Few assumptions  reduces the risk of those assumptions
being violated. Fewer and more realistic assumptions also make the particle 
filter more robust to unforeseen difficulties in FMRI data.

@@

\chapter{Future Work}
\label{sec:FutureWork}
% Enhancements/Dehancements 
%% prior
%%% flatten prior
%%% Larger variance
%% Detrending
%%% \% difference using spline?
%%% "smart knots"
%%% DC gain as a parameter
%%% linearizing
%% Experimental Changes
%%% Variations in stimuli
%%% longer timeseries or artificial
%%% Automatic estimation of measurement error.
\section{Algorithm Improvements}
\label{sec:Particle Filter Variations}
There are a few areas where future research may improve upon the current
work. The first is modifying the prior 
distribution. The prior distribution used in this work is listed in \autoref{tab:Prior}, 
and is based on the findings of Friston et al. \cite{Friston2000}. Unfortunately, that
result depended on a quadratic approximation (Volterra Series) which has
not been extensively tested (at least not in a published work). 
This is in additoin to the fact that the parameters are ill-defined, and
thus the parameters found in Friston et al. are most likely far too wide.
The prior is also
based completely on the BOLD output, which is why it is good at generating
estimates of the BOLD signal, but weak at estimating parameters. 
The best solution would be to have
in-vivo estimates of the actual parameters, although this
is unlikely to happen. Instead, better estimates could be found
using population studies with additional measurements as discussed in 
\autoref{sec:Sideways Measurements}. Better knowledge of the prior distribution
would make it possible to decrease variance of certain 
parameters, reducing the mobility of the parameters.

Another major area for improvement is removing drift. 
A large number of detrending methods were tested, but every method was in some way limited. 
One potentially viable method that could be utilized, given the right 
experimental design, is detrending based on intervals of low activity. This 
has the advantage that it wouldn't require an arbitrary
constant to be added to the pre-processed signal for the
BOLD model to fit properly.  The disadvantage of this approach is 
that it could hide long fall times by normalizing them
out. It also requires periodic breaks in the stimulus, 
which could reduce value of those samples for fitting purposes.

In terms of dealing with drift, linearizing is another possibility.
This would use the delta between measurements for fitting, rather than the 
direct value. This has the advantage of not requiring detrending and thus 
gives nearly raw data to the particle filter for processing. 
The effectiveness of this method depends directly on the type of stimulus.
In a test with rapid impulse stimuli, this could be extremely effective because
the DC level has minimal data; conversely prolonged flat levels make this less effective.
In tests I found that large drift/low white noise
type signals performed much better with a linearization approach.
For the stimulus sequence used in \autoref{sec:RealData}, the results tended to be worse
than using spline detrending.

@@

\section{Experimental Design Improvements}
\label{sec:Sideways Measurements}
One definite way of improving the results of the particle filter is through
additional measurements. While increasing the sample rate of FMRI scanners may
not be possible, simultaneous measurements of volume and flow is
possible, albeit at 9T in a cat \cite{Hu2009}. Just one of those measurements
though would be extremely powerful when incorporated into the BOLD
model. By adding another measurement, the
variance in the parameter estimates would significantly drop. The fact
that such a measurement would be closer to the true stimulus would make it
all the more powerful. 
A more conventional method of adding measurements is to 
simply perform longer FMRI tests. Although this wouldn't be 
groundbreaking, it would certainly increase the ability of the 
particle filter to develop the posterior distribution. 
It may also be possible to concatenate multiple FMRI tests together.
It should be possible to link the multiple runs simply through
the final parameter distribution from the previous run.

@@

% Future works
%% Using s as the input to other regions. 
%% Comparison of posterior of parameters
\section{Future Applications}
There are a number of advantages to the particle filter approach presented
here. In the past, FMRI data has been analyzed strictly for determining
correlation between a stimulus and response. With this new method
the correlation is simply a means to determining the joint posterior distribution
of the parameters. While only regions that correlate with the input will 
be calculable, this method exploits that correlation to constrain the 
prior distribution of the parameters. The resulting distribution, while difficult 
to visualize because of the high-dimensionality, could nevertheless be 
correlated with neural pathologies. In spite of the fact that the parameters
are under-determined, the final distribution is still a reduction in the
uncertainty of the parameters. The availability of a full posterior distribution
opens up many avenues for further inquiry, for instance
to compare normal vs. symptomatic
populations. This would be especially useful in patients whose symptoms
do not include structural changes in brain. This would effectively be
a way to differentiate the way the brain operates.

Because of limitations present in every imaging modality,
its becoming increasingly clear that combining data from multiple sources
will be necessary to push Neurology forward. In order to do so however, the
output of each source needs to fully represent what information that source
can provide. Combining the sources using Bayesian statistics is promising
yet often difficult because full probability distributions are hard to come by.
However, in this case, the particle filter provides a full posterior which
is extremely versatile. Therefore, future works will easily be able to
plug in data from multiple sources if they all output Bayesian posteriors.
For instance, if two different modalities have calculated the probability
distribution of neural efficiency, those two beliefs may be combined into
one conditional belief for the probability of neural efficiency.

>>>>

\chapter{Discussion}
\label{sec:Discussion}
\section{Review of Results}
% Overview of results
%% Parameters under-constrained
%% Output estimates good
There are two possible causes for the large variance in
parameter estimates. First, it could be that the particle filter is
incapable of learning the model. However, given the high quality
of BOLD estimates, this is almost certainly not the case. The other
possible explanation is that the given stimulus is not sufficiently
varied to bring out all the properties of the system. While it is
possible that this is the case, it is unlikely that any
stimulus sequence could differentiate all the parameters.
Its also extremely interesting that the histograms of parameters' mean
approach the width of the prior, even when the particle filter converges
properly to far thinner distributions. Its possible the prior
distribution is far wider than
it needs to be. Finally, the fact that \autoref{sec:VeryLongSim}, which
used randomly spaced impulses,
still had a high correlation further indicates that the parameters
are ill-defined.

Thus there is little doubt that the parameters of the BOLD model are under
constrained. While unsurprising given the sensitivity analysis by Deneux et al.,
it is a notable conclusion \cite{Deneux2006}. Other methods
that calculate a point estimate of the parameters, such as least squares
or Kalman filters (which uses the first two moments) are limited in their
capacity to estimate such parameters. While estimates of
the  BOLD signal may still be correct, the
underlying parameters and state variables cannot be described without using
a joint probability distribution function. In this sense, particle
filters represent an important step forward in BOLD parameter
estimation. Representing uncertainty with a mean
and variance is insufficient; so using a particle filter
or Bayesian estimate of the posterior is not simply an enhancement,
but a necessary precaution.

@@

As the results in \autoref{sec:RealData} show, the heat maps, especially
those of mutual information, closely resembled the results of SPM. While the activation
tests were more sensitive, there were some additional false positives,
though the problem is difficult to quantify. Regions with M.I. above
$0.15$ consistently fit the FMRI data well, and simulations showed that
the particle filter performed extremely well in the face of significant
noise. In sum, the estimated BOLD output remained consistent despite
large swings in parameters.

@@

%A practical benefit with the particle filter is that it is mathematically
%simple. An understanding of Bayesian statistics is all that is necessary to understand
%how the particle filter works.  Additionally very few assumptions
%are needed for the particle filter. Few assumptions  reduces the risk of those assumptions
%being violated. Fewer and more realistic assumptions also make the particle
%filter more robust to unforeseen difficulties in FMRI data.

@@

\chapter{Future Work}
\label{sec:FutureWork}
% Enhancements/Dehancements
%% prior
%%% flatten prior
%%% Larger variance
%% Detrending
%%% \% difference using spline?
%%% "smart knots"
%%% DC gain as a parameter
%%% linearizing
%% Experimental Changes
%%% Variations in stimuli
%%% longer timeseries or artificial
%%% Automatic estimation of measurement error.
\section{Algorithm Improvements}
\label{sec:Particle Filter Variations}
There are a few areas where future research may improve upon the current
work. The first is modifying the prior
distribution. The prior distribution used in this work is listed in \autoref{tab:Prior},
and is based on the findings of Friston et al. \cite{Friston2000}. Unfortunately, that
result depended on a quadratic approximation (Volterra Series) which has
not been extensively tested (at least not in a published work).
This is in addition to the fact that the parameters are ill-defined, and
thus the parameters found in Friston et al. are most likely far too wide.
The prior is also
based completely on the BOLD output, which is why it is good at generating
estimates of the BOLD signal, but weak at estimating parameters.
The best solution would be to have
in-vivo estimates of the actual parameters, although this
is unlikely to happen. Instead, better estimates could be found
using population studies with additional measurements as discussed in
\autoref{sec:Sideways Measurements}. Better knowledge of the prior distribution
would make it possible to decrease variance of certain
parameters, reducing the mobility of the parameters.

Another major area for improvement is removing drift.
A large number of detrending methods were tested, but every method was in some way limited.
One potentially viable method that could be utilized, given the right
experimental design, is detrending based on intervals of low activity. This
has the advantage that it wouldn't require an arbitrary
constant to be added to the pre-processed signal for the
BOLD model to fit properly. It would also minimize the actual
removed. The disadvantage of this approach is
that it could hide long fall times by normalizing them
out. It also requires periodic breaks in the stimulus,
which could reduce value of those samples for fitting purposes.

Linearizing is another possible method of dealing with drift.
This would use the delta between measurements for fitting, rather than the
direct value. This has the advantage of not requiring detrending and thus
gives nearly raw data to the particle filter for processing.
The effectiveness of this method depends directly on the type of stimulus.
In a test with rapid impulse stimuli, this could be extremely effective because
the DC signal has minimal data; conversely prolonged flat levels make this less effective.
In tests I found that large drift/low white noise
type signals performed much better with a linearization approach.
For the stimulus sequence used in \autoref{sec:RealData}, the results tended to be worse
than using spline detrending.

@@

\section{Experimental Design Improvements}
\label{sec:Sideways Measurements}
One definite way of improving the results of the particle filter is through
additional measurements. While increasing the sample rate of FMRI scanners may
not be possible, simultaneous measurements of volume and flow is
possible, albeit at 9T in a cat \cite{Hu2009}. Just one of those measurements
would be extremely powerful when incorporated into the BOLD
model. By adding another measurement, the
variance in the parameter estimates would significantly drop. The fact
that such a measurement would be closer to the true stimulus would make it
all the more useful.
A more conventional method of adding measurements is to
simply perform longer FMRI tests. Although this wouldn't be
ground breaking, it would certainly increase the ability of the
particle filter to develop the posterior distribution.
It may also be possible to concatenate multiple FMRI tests together
by using the parameter estimates from previous tests as the
prior for later tests.

@@

% Future works
%% Using s as the input to other regions.
%% Comparison of posterior of parameters
\section{Future Applications}
There are a number of advantages to the particle filter approach presented
here. In the past, FMRI data has been analyzed strictly for determining
correlation between a stimulus and response. With this new method
the correlation is simply a means to determining the joint posterior distribution
of the parameters. While only regions that correlate with the input will
be calculable, this method exploits that correlation to constrain the
prior distribution of the parameters. The resulting distribution, while difficult
to visualize because of the high-dimensionality, could nevertheless be
correlated with neural pathologies. In spite of the fact that the parameters
are under-determined, the final distribution is still a reduction in the
uncertainty of the parameters. The availability of a full posterior distribution
opens up many avenues for further inquiry, for instance
to compare normal vs. symptomatic
populations. This would be especially useful in patients whose symptoms
are caused by functional difference in the brain, rather than structural differences.
This would effectively be
a way to differentiate the way the brains operate.

Because of limitations present in every imaging modality,
its becoming increasingly clear that combining data from multiple sources
will be necessary to push Neurology forward. In order to do so however, the
output of each source needs to fully represent what information that source
can provide. Combining the sources using Bayesian statistics is promising
yet often difficult because full probability distributions are hard to come by.
However, in this case, the particle filter provides a full posterior which
is extremely versatile. Therefore, future works will easily be able to
plug in data from multiple sources if they all output Bayesian posteriors.
For instance, if two different modalities have calculated the probability
distribution of neural efficiency, those two beliefs may be combined into
one conditional belief for a new conditional probability of neural efficiency.
